### 2007-01-19 Documentation/part1.tex
*** empty log message ***

 for(i=1;i<4;i++) fprintf(stream,"1->\%i: \%g ",i,\\ \hspace*{4cm} glbProfileProbability(0,1,i,+1,50)); \\

@@

 minimum = glbAllocParams(); \\

@@

experiments.

@@

  fprintf(stream,"$\backslash$nchi2 for combination at minimum of Exp. 1:\\ \hspace*{0.5cm} \%g $\backslash$n",chi2); \\

@@

chi2 for combination at minimum of Exp. 1: 59.4879 \\

@@

We will in this part not go into details of the experiment
definition. The pre-defined experiment prototypes in the {\tt data}
subdirectory are summarized in \tabl{experiments}. They correspond
(except from minor modifications) to the experiments in the
respective references in the table. These files are installed to
the directory \verb+${prefix}/share/globes+ which usually defaults
to  \verb+/usr/local/share/globes+. It is useful to add this path
to the value of \verb+GLB_PATH+.
\index{norm}{glb files@{\tt glb}-files!installation}

@@

\begin{table}[tbp]
\begin{center}
\begin{tabular}{llp{7.5cm}c}
\hline
Experiment & Filename & Short description &  Ref. \\[0.2cm]
\hline 
\multicolumn{4}{l}{\underline{Superbeam experiments:}} \\[0.2cm]

@@

Double{\sc Chooz} & {\tt D-Chooz\_near.glb} & DoubleCHOOZ near 
detector, 5 years data taking & \cite{Huber:2006vr} \\ 
 & {\tt D-Chooz\_far.glb} & DoubleCHOOZ near detector, 5 years data taking & \\[0.2cm]

@@

low $\gamma$ & {\tt BB\_100.glb} & $\gamma=100$ CERN to Fr\'{e}jus baseline 
scenario, 4~yr $\nu$- and 4~yr $\bar{\nu}$-running & \cite{Campagne:2006yx} \\[0.2cm]

medium $\gamma$ & {\tt BB\_350.glb} & $\gamma=350$, refurbished SPS scenario, 4~yr $\nu$- and
4~yr $\bar{\nu}$-running & \cite{Burguet-Castell:2005pa} \\[0.2cm]

Variable & {\tt BBvar\_WC.glb} & Variable $\beta$-beam, 4~yr $\nu$- and 
4~yr $\bar{\nu}$-running, {\sf AEDL}-Variables {\tt gammafactor}, 
{\tt EXP\_FACTOR} and {\tt baselinefactor}, 500~kt WC detector & \cite{Huber:2005jk} \\[0.2cm]

	& {\tt BBvar\_TASD.glb} & like {\tt BBvar\_WC.glb} but with a 50~kt TASD detector  
& \cite{Huber:2005jk} \\[0.2cm]

@@

Variable  & {\tt NFvar.glb} & Variable neutrino factory, 4~yr 
$\nu$- and 4~yr $\bar{\nu}$-running, {\sf AEDL}-Variables {\tt emax} and {\tt BASELINE}, disappearance channels
{\em without} CID & \cite{Huber:2002mx,Huber:2006wb}  \\[0.2cm]

Gold + Silver & {\tt NF\_GoldSilver.glb} & as {\tt NFvar.glb} 
plus 5~kt ECC detector for Silver Channel
measurement & \cite{Huber:2002mx,Huber:2006wb,Autiero:2003fu} \\[0.2cm]

Hybrid detector & {\tt NF\_hR\_lT.glb} & like {\tt NFvar.glb} 
with lower threshold and higher energy resolution & \cite{Huber:2002mx,Huber:2006wb} \\[0.2cm]

\hline
\end{tabular}
\end{center}
\mycaption{\label{tab:experiments} \index{norm}{Experiment files (table)}
\index{norm}{glb files@{\tt glb}-files} 
Pre-defined experiment prototypes, their filenames (to be used in \GLB{InitExperiment}), 
their short descriptions, and the references in which they are originally used and 
discussed (except from minor modifications, such as a different implementation of 
the energy threshold function). Note that some of these experiments are outdated 
in terms of integrated luminosities, baseline, fluxes, efficiencies, or other factors. 
In any case these file are installed along with \GLOBES.}
\end{table}

@@

\index{norm}{Experiment!number of}
In principle, the \GLOBES\ user interface can currently handle up to 32 of different long-baseline experiments simultaneously, where the number
of existing experiment definition files can, of course, be unlimited. This means that their $\Delta \chi^2$-values are added {\em after} the minimization over the systematics parameters, and {\em before} any minimization over the oscillation parameters. Note that each experiment
assumes a specific matter density profile, which means
that it makes sense to simulate different operation modes within one
experiment definition, and physically different baselines, in different
definitions. For details of the rate computation and
simulation techniques, we refer at this place to \Part~\ref{part:2}. Though
 the simplest case of simulating one experiment may be most often used, 
 using several experiments is useful in many cases. For example, combinations of experiments can be tested for
complementarity and competitiveness by equal means within one program.
In general, many \GLOBES\ functions take the experiment number as
a parameter, which runs from $0$ to \GLB{\_num\_of\_exps}-1 in the order of their initialization in the program.\footnote{Note that
the global variable {\tt glb\_num\_of\_exps} must not be modified by the
user.} In addition, using the parameter value \GLBC{GLB\_ALL} as
experiment number (for example, in the {\tt glbChi...} functions) initiates a combined analysis of all loaded experiments.
\index{norm}{GLB ALL@{\tt GLB\_ALL}}
\index{constants}{{\tt GLB\_ALL}}

@@

\index{norm}{Path resolution}
A list of pre-defined experiment prototypes, their filenames, their short descriptions, and the references of their definitions can be found in \tabl{experiments}. If the program cannot find these files, or some of them are syntactically not correct, it will break at this place.

@@

Compared to existing experiments, which use real data, future experiments
use simulated data. Thus, the {\em true parameter values} and their results in 
form of the reference event rate vectors are simulated. After setting the 
true parameter values, the {\em fit parameter values} can be varied in order to 
obtain information on the measurement performance for the given set of true 
parameter values. Therefore, it is often useful to show the results of a 
future measurement as function of the true parameter values for which the 
reference rate vectors are computed -- at least within the currently allowed 
ranges. The true parameter values for the vacuum neutrino oscillation 
parameters have to be set by the function \GLB{SetOscillationParameters}, 
and the reference rate vector, \ie\ the data, has to be computed by a call 
to \GLB{SetRates}.
This has to be done {\em before} any evaluation 
function is used and  {\em after} the experiments have been initialized and 
also the experiment 
parameters have been adjusted which could change the rates (such as baseline
 or target mass). This means that
after any change of an experiment parameter, \GLB{SetRates} has to be 
called. Matter effects are automatically included as specified in the 
experiment definition. We have the following functions to assign and
read out the vacuum oscillation parameters:
\begin{function}
\GLBNS{SetOscillationParameters}
{\tt int glbSetOscillationParameters(const glb\_params in)} sets the
vacuum oscillation parameters to the ones in the vector {\tt in}.
\end{function}
\begin{function}
\GLBNS{GetOscillationParameters}
{\tt int glbGetOscillationParameters(glb\_params out)} returns the
vacuum oscillation parameters in the vector {\tt out}. The result of
the function is $0$ if the call was successful.
\end{function}
The reference rate vector is then computed with:
\begin{function}
\GLBNS{SetRates}
{\tt void glbSetRates()} computes the reference rate vector for the neutrino oscillation parameters set by {\tt glbSetOscillationParameters}. 
\end{function}
A complete example for a minimal \GLOBES\ program can be found on
Page~\pageref{ex:c}.

@@

In order to keep track of the used version of \GLOBES , the software
provides a number of functions to check the \GLOBES\ and experiment
versions. It is up to the user to implement mechanisms into the
program and \AEDL\ files to check whether
\begin{itemize}
\item
 The program should only run with this specific version of \GLOBES .
\item
 The program can only run up to a certain \GLOBES\ version.
\item
 The program can only run with a minimum version of \GLOBES .
\item
 The program and AEDL file versions are compatible.
\end{itemize}
However, note that GLOBES~3.0 and higher requires that the \GLOBES\ version be at least as new as the
version the AEDL file was written for.
%
The functions in \GLOBES\ for version control are:
\begin{function}
\GLBNS{TestReleaseVersion}
{\tt int glbTestReleaseVersion(const char *version)} returns $0$ if
the {\tt version} string of the format ``X.Y.Z'' is exactly the
used \GLOBES\ version, $1$ if it is older, and $-1$ if it is newer. 
\end{function}
% UNDOCUMENTED FUNCTION, SINCE VERY DIFICULT TO EXPLAIN DIFFERENCE:
% \begin{function}
% \GLBNS{TestLibraryVersion}
% {\tt int glbTestLibraryVersion(const char *version)} returns $0$ if
% the {\tt version} string of the format ``X.Y.Z'' is exactly the
% used \GLOBES\ library version, $1$ if it is older, and $-1$ if it is newer.
% \end{function}
% Note that the library and \GLOBES\ versions are not the same.

\begin{function}
\GLBNS{VersionOfExperiment} 
{\tt const char* glbVersionOfExperiment(int experiment)} returns
the version string of the experiment number {\tt experiment} (set by
{\tt \$version} in AEDL).
The version string is allocated within the experiment structure, which means
that it cannot be altered and must not be freed by the
user.
\end{function}
%
A useful function to debug \GLOBES\ source code is 
\begin{function}
\GLBNS{SetVerbosityLevel}
{\tt int glbSetVerbosityLevel(int level)} sets the verbosity level
for \GLOBES\ messages.
\end{function}
The following verbosity levels are currently used:
\begin{description}
\item[0] Display no messages
\item[1] (Standard) Display error messages
\item[2] Display warnings
\item[3] Display file access history
\item[4] Display search paths
\end{description}
Note that always messages with the chosen verbosity level and lower
are displayed.

@@

Calculating a $\chi^2$-value with or without systematics, but no correlations and degeneracies, is the simplest and fastest possibility to obtain high-level information on an experiment. In general, \GLOBES\ uses the six independent oscillation parameters $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, $\deltacp$, $\sdm$, $\ldm$, as well as the matter density 
scaling factor $\hat{\rho}$ of each experiment.\index{norm}{Matter density!
scaling factor} Thus, there are six plus the number of experiments parameters determining the rate vectors. Using the matter density scaling factors  in addition to the oscillation parameters will allow the simulation of the correlations with matter density uncertainties: In this
approach, the matter density profile normalization $\hat{\rho}$ can 
be treated
as parameter to be measured by the experiment, where an external precision
given by observations is imposed (typically up to $5\%$). 
\index{norm}{Matter density!uncertainty}
 For this section, it is important to keep in mind
that there are more parameters than just the oscillation parameters
determining the simple $\chi^2$. However, as we have described in the
last section, the mechanism for the matter density scaling factors
is hidden in the definition of \GLB{\_params}: Each of the scaling
factors is initially set to $1.0$. Therefore, for the calculation of
$\chi^2$ with systematics only, we do not have to care about the
matter density scaling factors.

@@

The resulting data can then be plotted as a contour plot (2 d.o.f.):
\begin{center}
\colorbox{white}{\includegraphics[width=8cm]{correx}}
\end{center}

@@

Keeping all oscillation parameters and matter density scaling factors fixed,
 one can use the following builtin functions to obtain the total $\chi^2$ of all 
 specified oscillation channels including systematics:
\begin{function} 
\GLBNS{ChiSys}
{\tt double glbChiSys(const glb\_params in,int exp, int rule)} returns
the $\chi^2$ for the (fixed) oscillation parameters {\tt in}, the
experiment number {\tt exp}, and the rule number {\tt rule}. For all
experiments or rules, use \GLBC{GLB\_ALL} as parameter value.
\end{function}
Note that the result of {\tt glbChiSys} for all experiments or rules
corresponds to the sum of all of the individual {\tt glbChiSys} calls. 
This equality will not hold for the minimizers in the next sections anymore. 
 An example how to use  {\tt glbChiSys} can be found on page~\pageref{ex:corrth13dcp}.

\index{norm}{Systematics} \index{norm}{Pull method}
The treatment of systematics in \GLOBES\ is performed by the so-called
{\em pull method} with the help of auxiliary systematics parameters.
They are taken to be completely uncorrelated among different rules,
and treated with simple Gau\ss ian statistics.
In general, a rule is a prescription for summing up experimentally indistinguishable
signal and background events from different oscillation channels.
For more details on the rule concept, see \Part~\ref{part:2} of this manual,
and for the treatment of systematics, see \Sec~\ref{sec:rules}.

 One example for a systematics parameter is the signal normalization error, \ie, an error on the overall normalization of the signal. For illustration, we assume that the signal event rate in the $i$th bin $s_i^0$ of one oscillation channel is altered by the overall normalization auxiliary parameter\index{norm}{Auxiliary parameter} of this channel, \ie , 
\be
 s_i = s_i(n_s) = s_i^0 \cdot (1 + n_s),
\ee
where $n_s$ is the signal normalization parameter. The total number of events in the $i$th bin $x_i$ also includes the background event rates $b_i$, \ie, $x_i = s_i + b_i$, which may have their own systematics parameters.
In order to implement an overall signal normalization error $\sigma_{n_s}$,  the $\chi^2$, which includes all event rates $x_i$ of all bins, is minimized over the auxiliary parameter $n_s$:
\be
 \hat{\chi^2} = \underset{n_s}{\mathrm{min}} \left(  \chi^2(n_s, \hdots) + \frac{(n_s)^2}{\sigma_{n_s}^2} \right).
\ee 
This minimization is done independently for all auxiliary parameters of the rule. The total $\chi^2$ for the considered experiment is finally obtained by repeating this procedure for all rules and adding their $\chi^2$-values. In general, the situation is more complicated because of the usage of many systematical errors. More details about systematics parameters and the definition of signal, background, and oscillation channels can be found in \Sec~\ref{sec:rules}, too.

@@

\section{User-defined systematics calculation}
\index{norm}{Systematics!user-defined}
\label{sec:userchi}

@@

For some experimental setups, the built-in systematics functions are not sufficient
because they cannot handle systematical errors that are correlated between different
parts of a multi-detector setup. Therefore \GLOBES\ allows the user
to override the default $\chi^2$ calculation. The reader who wants to use this
feature should be familiar with \Secs~\ref{sec:event_rates},~\ref{sec:channel}, and~\ref{sec:rules}
of the manual. Note that this is an advanced topic, which requires \GLOBES\ 3.0 or
higher. It can be omitted in a first reading of the manuscript.

User-defined systematics is, compared to the built-in systematics, a compound between
the AEDL definition and some source code defining the behavior. In the simplest case,
the AEDL file defines user-defined systematics for a specific rule, gives it a 
name/identifier, and defines the systematical errors. The source code has then to match this AEDL
definition -- therefore, make sure to use only unique identifiers. 
Ones needs to register the user-defined systematics in the 
source code (typically after {\tt glbInit}) with
\GLBNS{DefineChiFunction}
{\tt int glbDefineChiFunction(glb\_chi\_function chi\_func, int dim, const char *name, void* user\_data)}
 tells \GLOBES\ to register the user-defined systematics $\chi^2$ function {\tt chi\_func} 
identified by the string {\tt name} with {\tt dim} systematics parameters. The parameter {\tt user\_data}
is an arbitrary pointer being transferred to {\tt chi\_func}. It can, for instance, be used to avoid
global variables. Returns zero, if successful.
The user-defined systematics function itself has to have the pre-defined type 
\GLBNS{\_chi\_function}
{\tt double (*glb\_chi\_function)(int exp, int rule, int dim, double *params, double *errors, void* user\_data)}.
In this function type, {\tt exp} is the experiment number, {\tt rule} is the rule number, {\tt dim} is the number of systematics parameters, {\tt params} is an array of the systematics auxiliary parameters themselves, and {\tt errors} is an array with the systematical errors. The parameter {\tt user\_data} is set as defined with
{\tt glbDefineChiFunction}.
Note the the central values for {\tt params} are $0$, which means that $0$ corresponds to the un-modified rates. The array indices run from $0$ to {\tt dim}-1.
Note that this function will be called many times by the \GLOBES\ minimizers.
Therefore the function should be as efficient as possible.
%JK - THIS IS INCORRECT: ONE MINIMIZATION OVER, SAY, 8 PARAMETERS IS FASTER THAN
%2 MINIMIZATIONS OVER 4 PARAMETERS EACH-
%In many cases, it is also recommendable to use different functions for different
%rules, since the fewer auxiliary parameters in the minimization, the faster the routine will be.
In addition, note that complicated systematics with many parameters may introduce complicated (unknown) topologies for the minimization, which means that the minimizers may
end up in a local minimum instead of the global minimum. \GLOBES\ provides the function \GLB{SetSysStartingValuesList} to change the starting values of the minimizer in the case of
 convergence problems (see below).
A typical implementation may look like the following code, where ``{\tt chiDCNorm}'' is the identifier
assigned in the AEDL definition:

>>>>

 for(i=1;i<4;i++) fprintf(stream,"1->\%i: \%g ",i,\\ \hspace*{4cm} glbProfileProbability(0,1,i,+1,50)); 

@@

 minimum = glbAllocParams(); 

@@

two baselines:

@@

  fprintf(stream,"$\backslash$nchi2 for combination at minimum of L=3000km:\\ \hspace*{0.5cm} \%g $\backslash$n",chi2); \\

@@

chi2 for combination at minimum of L=3000km: 59.4879 \\

@@

We will in this part not go into details of the experiment
definition. The pre-defined experiment prototypes in the {\tt data}
subdirectory are summarized in \tabl{experiments} and described in
\App~\ref{app:aedlfiles}. They correspond
(except from minor modifications) to the experiments in the
respective references in the table. These files are installed to
the directory \verb+${prefix}/share/globes+ which usually defaults
to  \verb+/usr/local/share/globes+. It is useful to add this path
to the value of \verb+GLB_PATH+.
\index{norm}{glb files@{\tt glb}-files!installation}

@@

\begin{table}[tbp]
\begin{center}
{\small
\begin{tabular}{llp{7.1cm}c}
\hline
Experiment & Filename & Short description &  \Ref \\
\hline 
\multicolumn{4}{l}{\underline{Superbeam experiments:}} \\[0.2cm]

@@

Double{\sc Chooz} & {\tt D-Chooz\_near.glb} & Double Chooz near 
detector, 5 years data taking & \cite{Huber:2006vr} \\ 
 & {\tt D-Chooz\_far.glb} & Double Chooz far detector, 5 years data taking & \\[0.2cm]

@@

Low $\gamma$ & {\tt BB\_100.glb} & $\gamma=100$ CERN to Fr\'{e}jus baseline 
scenario, 4~yr $\nu$- and 4~yr $\bar{\nu}$-running & \cite{Campagne:2006yx} \\[0.2cm]

Medium $\gamma$ & {\tt BB\_350.glb} & $\gamma=350$, refurbished SPS scenario, 4~yr $\nu$- and
4~yr $\bar{\nu}$-running & \cite{Burguet-Castell:2005pa} \\[0.2cm]

Variable $\gamma$ & {\tt BBvar\_WC.glb} & Variable $\beta$-beam, 4~yr $\nu$- and 
4~yr $\bar{\nu}$-running, {\sf AEDL}-Variables {\tt gammafactor}, 
{\tt EXP\_FACTOR} and {\tt baselinefactor}, 500~kt WC detector & \cite{Huber:2005jk} \\[0.2cm]

	& {\tt BBvar\_TASD.glb} & As {\tt BBvar\_WC.glb}, but with a 50~kt TASD
& \cite{Huber:2005jk} \\[0.2cm]

@@

Variable $E_\mu$ & {\tt NFvar.glb} & Variable $E_\mu$ neutrino factory, 4~yr 
$\nu$- and 4~yr $\bar{\nu}$-running, {\sf AEDL}-Variables {\tt emax} and {\tt BASELINE}, disapp. channels
{\em without} CID & \cite{Huber:2002mx,Huber:2006wb}  \\[0.2cm]

Gold + Silver & {\tt NF\_GoldSilver.glb} & As {\tt NFvar.glb}, 
plus 5~kt ECC detector for Silver Channel
measurement & \cite{Huber:2002mx,Huber:2006wb,Autiero:2003fu} \\[0.2cm]

Hybrid det. & {\tt NF\_hR\_lT.glb} & As {\tt NFvar.glb}, 
but lower threshold and higher energy resolution & \cite{Huber:2002mx,Huber:2006wb} \\[0.2cm]

\hline
\end{tabular}
} % small
\end{center}
\mycaption{\label{tab:experiments} 
%
\index{norm}{Experiment files (table)}
\index{norm}{glb files@{\tt glb}-files} 
% DOES NOT WORK. FOR SOME REASON, THERE ARE DOUBLE ENTRIES IN THE INDEX, IF THE FOLLOWING CODE IS USED [WW]:
% \index{aedl}{T2K.glb@{\tt T2K.glb}}
% \index{aedl}{T2HK.glb@{\tt T2HK.glb}}
% \index{aedl}{NOvA.glb@{\tt NOvA.glb}}
% \index{aedl}{SPL.glb@{\tt SPL.glb}}
% \index{aedl}{Reactor1.glb@{\tt Reactor1.glb}}
% \index{aedl}{Reactor2.glb@{\tt Reactor2.glb}}
% \index{aedl}{D-Chooznear.glb@{\tt D-Chooz\_near.glb}}
% \index{aedl}{D-Choozfar.glb@{\tt D-Chooz\_far.glb}}
% \index{aedl}{BB100.glb@{\tt BB\_100.glb}}
% \index{aedl}{BB350.glb@{\tt BB\_350.glb}}
% \index{aedl}{BBvarWC.glb@{\tt BBvar\_WC.glb}}
% \index{aedl}{BBvarTASD.glb@{\tt BBvar\_TASD.glb}}
% \index{aedl}{NFstandard.glb@{\tt NFstandard.glb}}
% \index{aedl}{NFvar.glb@{\tt NFvar.glb}}
% \index{aedl}{NFGoldSilver.glb@{\tt NF\_GoldSilver.glb}}
% \index{aedl}{NFhRlT.glb@{\tt NF\_hR\_lT.glb}}
%
Pre-defined experiment prototypes, their filenames (to be used in \GLB{InitExperiment}), 
their short descriptions, and the references in which they are originally used and 
discussed (except from minor modifications, such as a different implementation of 
the energy threshold function). Note that some of these experiments may want to be
adjusted in terms of integrated luminosity, baseline, flux, efficiencies, or other factors. 
In any case these file are installed along with \GLOBES . For more details, see \App~\ref{app:aedlfiles}.}
\end{table}

@@

\index{norm}{Experiment!number of}
In principle, the \GLOBES\ user interface can currently handle up to 32 of different long-baseline experiments simultaneously, where the number
of existing experiment definition files can, of course, be unlimited.  Note that each experiment
assumes a specific matter density profile, which means
that it makes sense to simulate different operation modes within one
experiment definition, and physically different baselines, in different
definitions. For details of the rate computation and
simulation techniques, we refer at this place to \Part~\ref{part:2}. Though
 the simplest case of simulating one experiment may be most often used, 
 using several experiments is useful in many cases. For example, combinations of experiments can be tested for
complementarity and competitiveness by equal means within one program.
In general, many \GLOBES\ functions take the experiment number as
a parameter, which runs from $0$ to \GLB{\_num\_of\_exps}-1 in the order of their initialization in the program.\footnote{Note that
the global variable {\tt glb\_num\_of\_exps} must not be modified by the
user.} In addition, using the parameter value \GLBC{GLB\_ALL} as
experiment number (for example, in the {\tt glbChi...} functions) initiates a combined analysis of all loaded experiments.
\index{norm}{GLB ALL@{\tt GLB\_ALL}}
\index{constants}{{\tt GLB\_ALL}}

@@

\index{norm}{Path resolution}
A list of pre-defined experiment prototypes, their filenames, their short descriptions, and the references of their definitions can be found in \tabl{experiments}; for more details, see \App~\ref{app:aedlfiles}. If the program cannot find these files, or some of them are syntactically not correct, it will break at this place.

@@

Compared to existing experiments, which use real data, future experiments
use simulated data. Thus, the {\em true parameter values} (or simulated
parameter values) are used to calculate the reference event rate vectors
corresponding to the simulated experiment result. After setting the 
true parameter values, the {\em fit parameter values} can be varied in order to 
obtain information on the measurement performance for the given set of true 
parameter values. Therefore, it is often useful to show the results of a 
future measurement as function of the true parameter values for which the 
reference rate vectors are computed -- at least within the currently allowed 
ranges. The true parameter values for the vacuum neutrino oscillation 
parameters have to be set by the function \GLB{SetOscillationParameters}, 
and the reference rate vector, \ie\ the data, has to be computed by a call 
to \GLB{SetRates}.
This has to be done {\em before} any evaluation 
function is used and  {\em after} the experiments have been initialized and 
also the experiment 
parameters have been adjusted which could change the rates (such as baseline
 or target mass). This means that
after any change of an experiment parameter, \GLB{SetRates} has to be 
called. Matter effects are automatically included as specified in the 
experiment definition. We have the following functions to assign and
read out the vacuum oscillation parameters:
\begin{function}
\GLBNS{SetOscillationParameters}
{\tt int glbSetOscillationParameters(const glb\_params in)} sets the
vacuum oscillation parameters to the ones in the vector {\tt in}.
\end{function}
\begin{function}
\GLBNS{GetOscillationParameters}
{\tt int glbGetOscillationParameters(glb\_params out)} returns the
vacuum oscillation parameters in the vector {\tt out}. The result of
the function is $0$ if the call was successful.
\end{function}
The reference rate vector is then computed with:
\begin{function}
\GLBNS{SetRates}
{\tt void glbSetRates()} computes the reference rate vector for the neutrino oscillation parameters set by {\tt glbSetOscillationParameters}. 
\end{function}
A complete example for a minimal \GLOBES\ program can be found on
Page~\pageref{ex:c}.

@@

In order to keep track of the used version of \GLOBES , the software
provides a number of functions to check the \GLOBES\ and experiment
versions. It is up to the user to implement mechanisms into the
program and \AEDL\ files to check whether
\begin{itemize}
\item
 The program should only run with this specific version of \GLOBES .
\item
 The program can only run up to a certain \GLOBES\ version.
\item
 The program can only run with a minimum version of \GLOBES .
\item
 The program and \AEDL\ file versions are compatible.
\end{itemize}
However, note that \GLOBES ~3.0 and higher requires that the \GLOBES\ version be at least as new as the
version the \AEDL\ file was written for.
%
The functions in \GLOBES\ for version control are:
\begin{function}
\GLBNS{TestReleaseVersion}
{\tt int glbTestReleaseVersion(const char *version)} returns $0$ if
the {\tt version} string of the format ``X.Y.Z'' is exactly the
used \GLOBES\ version, $1$ if it is older, and $-1$ if it is newer. 
\end{function}
% UNDOCUMENTED FUNCTION, SINCE VERY DIFICULT TO EXPLAIN DIFFERENCE:
% \begin{function}
% \GLBNS{TestLibraryVersion}
% {\tt int glbTestLibraryVersion(const char *version)} returns $0$ if
% the {\tt version} string of the format ``X.Y.Z'' is exactly the
% used \GLOBES\ library version, $1$ if it is older, and $-1$ if it is newer.
% \end{function}
% Note that the library and \GLOBES\ versions are not the same.

\begin{function}
\GLBNS{VersionOfExperiment} 
{\tt const char* glbVersionOfExperiment(int experiment)} returns
the version string of the experiment number {\tt experiment} (set by
{\tt \$version} in \AEDL ).
The version string is allocated within the experiment structure, which means
that it cannot be altered and must not be freed by the
user.
\end{function}
%
A useful function to debug \GLOBES\ source code is 
\begin{function}
\GLBNS{SetVerbosityLevel}
{\tt int glbSetVerbosityLevel(int level)} sets the verbosity level
for \GLOBES\ messages.
\end{function}
The following verbosity levels are currently used:
\begin{description}
\item[0] Display no messages
\item[1] (Standard) Display error messages
\item[2] Display warnings
\item[3] Display file access history
\item[4] Display search paths
\end{description}
Note that always messages with the chosen verbosity level and lower
are displayed.

@@

Calculating a $\chi^2$-value including systematics, but without correlations and degeneracies, is the simplest and fastest possibility to obtain high-level information on an experiment. Here we describe the use of built-in
systematics, and the rather advanced topic of user-defined systematics. Note that the matter density 
is treated as an oscillation parameter in \GLOBES , which means that it is not dealt with at the
systematics level.

@@

The resulting data can then be plotted as a contour plot (2 d.o.f.):
\begin{center}
\colorbox{white}{\includegraphics[width=7.7cm]{correx}}
\end{center}

@@

Keeping all oscillation parameters and matter density scaling factors fixed,
 one can use the following builtin functions to obtain the total $\chi^2$ of all 
 specified oscillation channels including systematics:
\begin{function} 
\GLBNS{ChiSys}
{\tt double glbChiSys(const glb\_params in,int exp, int rule)} returns
the $\chi^2$ for the (fixed) oscillation parameters {\tt in}, the
experiment number {\tt exp}, and the rule number {\tt rule}. For all
experiments or rules, use \GLBC{GLB\_ALL} as parameter value.
\end{function}
Note that the result of {\tt glbChiSys} for all experiments or rules
corresponds to the sum of all of the individual {\tt glbChiSys} calls. 
This equality will not hold for the minimizers in the next chapters anymore. 
 An example how to use  {\tt glbChiSys} can be found on page~\pageref{ex:corrth13dcp}.

\index{norm}{Systematics} \index{norm}{Pull method}
The treatment of systematics in \GLOBES\ is performed by the so-called
{\em pull method} with the help of nuisance systematics parameters.
They are taken to be completely uncorrelated among different rules,
and treated with simple Gau\ss ian statistics.
In general, a rule is a prescription for summing up experimentally indistinguishable
signal and background events from different oscillation channels.
For more details on the rule concept, see \Part~\ref{part:2} of this manual,
and for the treatment of systematics, see \Sec~\ref{sec:rules}.

 One example for a systematics parameter is the signal normalization error, \ie, an error on the overall normalization of the signal. For illustration, we assume that the signal event rate in the $i$th bin $s_i^0$ of one oscillation channel is altered by the overall normalization nuisance parameter\index{norm}{Nuisance parameter} of this channel, \ie , 
\be
 s_i = s_i(n_s) = s_i^0 \cdot (1 + n_s),
\ee
where $n_s$ is the signal normalization parameter. The total number of events in the $i$th bin $x_i$ also includes the background event rates $b_i$, \ie, $x_i = s_i + b_i$, which may have their own systematics parameters.
In order to implement an overall signal normalization error $\sigma_{n_s}$,  the $\chi^2$, which includes all event rates $x_i$ of all bins, is minimized over the nuisance parameter $n_s$:
\be
 \hat{\chi^2} = \underset{n_s}{\mathrm{min}} \left(  \chi^2(n_s, \hdots) + \frac{(n_s)^2}{\sigma_{n_s}^2} \right).
\ee 
This minimization is done independently for all nuisance parameters of the rule. The total $\chi^2$ for the considered experiment is finally obtained by repeating this procedure for all rules and adding their $\chi^2$-values. In general, the situation is more complicated because of the usage of many systematical errors. More details about systematics parameters and the definition of signal, background, and oscillation channels can be found in \Sec~\ref{sec:rules}, too.

@@

\section{User-defined systematics calculation$^*$}
\index{norm}{Systematics!user-defined}
\label{sec:userchi}

@@

For some experimental setups, the built-in systematics functions are not sufficient
because they cannot handle systematical errors that are correlated between different
parts of a multi-detector setup. Therefore \GLOBES\ allows the user
to override the default $\chi^2$ calculation. The reader who wants to use this
feature should be familiar with \Secs~\ref{sec:event_rates},~\ref{sec:systematics},~\ref{sec:channel}, and~\ref{sec:rules}
of the manual. Note that this is an advanced topic, which requires \GLOBES\ 3.0 or
higher. It can be omitted in a first reading of the manuscript.

User-defined systematics is, compared to the built-in systematics, a compound between
the \AEDL\ definition and some source code defining the behavior. In the simplest case,
the \AEDL\ file defines user-defined systematics for a specific rule, gives it a 
name/identifier, and defines the systematical errors. The source code has then to match this \AEDL\
definition -- therefore, make sure to use only unique identifiers. 
Ones needs to register the user-defined systematics in the 
source code (typically after {\tt glbInit}) with
\begin{function}
\GLBNS{DefineChiFunction}
{\tt int glbDefineChiFunction(glb\_chi\_function chi\_func, int dim, const char *name, void* user\_data)}
 tells \GLOBES\ to register the user-defined systematics $\chi^2$ function {\tt chi\_func} 
identified by the string {\tt name} with {\tt dim} systematics parameters. The parameter {\tt user\_data}
is an arbitrary pointer being transferred to {\tt chi\_func}. It can, for instance, be used to avoid
global variables. Returns zero, if successful.
\end{function}
The user-defined systematics function itself has to have the pre-defined type 
\begin{function}
\GLBNS{\_chi\_function}
{\tt double (*glb\_chi\_function)(int exp, int rule, int dim, double *params, double *errors, void* user\_data)}.
In this function type, {\tt exp} is the experiment number, {\tt rule} is the rule number, {\tt dim} is the number of systematics parameters, {\tt params} is an array of the systematics nuisance parameters themselves, and {\tt errors} is an array with the systematical errors. The parameter {\tt user\_data} is set as defined with
{\tt glbDefineChiFunction}.
Note the the central values for {\tt params} are $0$, which means that $0$ corresponds to the un-modified rates. The array indices run from $0$ to {\tt dim}-1.
\end{function}
Note that this function will be called many times by the \GLOBES\ minimizers.
Therefore the function should be as efficient as possible.
%JK - THIS IS INCORRECT: ONE MINIMIZATION OVER, SAY, 8 PARAMETERS IS FASTER THAN
%2 MINIMIZATIONS OVER 4 PARAMETERS EACH-
%In many cases, it is also recommendable to use different functions for different
%rules, since the fewer auxiliary parameters in the minimization, the faster the routine will be.
In addition, note that complicated systematics with many parameters may introduce complicated (unknown) topologies for the minimization, which means that the minimizers may
end up in a local minimum instead of the global minimum. \GLOBES\ provides the function \GLB{SetSysStartingValuesList} to change the starting values of the minimizer in the case of
 convergence problems (see below).
A typical implementation may look like the following code, where ``{\tt chiDCNorm}'' is the identifier
assigned in the \AEDL\ definition:
\begin{quote}
{\tt
  double chiDCNorm(int exp, int rule, int dim, double *x, \\
  \hspace*{0.5cm} double *errors) \\
  \{ \\
   \hspace*{0.5cm} double chi2 = 0.0; \\
   \hspace*{0.5cm} int i; \\
   \\
   \hspace*{0.5cm} ... /* Some code to calculate the chi2 */ \\
   \\
    \hspace*{0.5cm} /* Here the systematics priors (penalties) are added: */ \\
    \hspace*{0.5cm} for(i=0;i<dim;i++) chi2 += (x[i]*x[i])/(errors[i]*errors[i]);
    \hspace*{0.5cm} return chi2; \\
  \} \\
   \\
  int main(int argc,char*argv[]) \\
 \{ \\ 
  \hspace*{0.5cm} ... \\
  \hspace*{0.5cm} glbInit(agrv[0]); \\
  \hspace*{0.5cm} glbDefineChiFunction(\&chiDCNorm,5,"chiDCNorm"); \\
  \hspace*{0.5cm} ... \\
 \} \\
}
\end{quote}

During running time, the systematics can be changed with 
\GLB{SetChiFunction} and \GLB{GetChiFunction} as described in \Sec~\ref{sec:systematics}.
In addition, especially useful in the context of user-defined systematics, a pointer to the systematics function can be returned either by name, or by experiment and rule selection:
\begin{function}
\GLBNS{GetChiFunctionPtr}
{\tt glb\_chi\_function glbGetChiFunctionPtr(const char *name)}
returns a pointer to the systematics $\chi^2$ function with a specified name {\tt name}.
\end{function}
\begin{function}
\GLBNS{GetChiFunctionPtrInExperiment}
{\tt glb\_chi\_function glbGetChiFunctionPtrInExperiment(int exp, int rule, int on\_off)}
returns a pointer to the systematics $\chi^2$ function of experiment {\tt exp} and rule {\tt rule}.
Systematics on or off can be accessed by {\tt on\_off}.
\end{function}
